---
title: "STAT 184 – Activity 14: A First QMD File"
author: "Devin Charles"
date: "November 10, 2025"
format: 
  pdf: 
    code-fold: false
    code-overflow: wrap
    code-line-wrap: true
execute:
  echo: true
  warning: false
  message: false
---

# Armed Forces Data Wrangling Redux

## Question 3: Data Wrangling Code for Armed Forces Data

Here is where I tidied my code from activity #10, but as a slightly revised version.

```{r}
library(dplyr)
library(tidyr)
library(knitr)

#load the data
armed_forces <- read.csv("active_duty_individual_8.csv", 
                         stringsAsFactors = FALSE)

#rank mapping table
rank_map <- tribble(
  ~Pay_Grade, ~Army, ~Navy, ~MarineCorps, ~AirForce, ~SpaceForce, ~CoastGuard,
  "E1", "Private", "Seaman Recruit", "Private", "Airman Basic", "Specialist 1",
  "Seaman Recruit",
  "E2", "Private", "Seaman Apprentice", "Private First Class", "Airman", 
  "Specialist 2", "Seaman Apprentice",
  "E3", "Private First Class", "Seaman", "Lance Corporal", 
  "Airman First Class", "Specialist 3", "Seaman",
  "E4", "Corporal OR Specialist", "Petty Officer Third Class", "Corporal", 
  "Senior Airman", "Specialist 4", "Petty Officer Third Class",
  "E5", "Sergeant", "Petty Officer Second Class", "Sergeant", "Staff Sergeant", 
  "Sergeant", "Petty Officer Second Class",
  "E6", "Staff Sergeant", "Petty Officer First Class", "Staff Sergeant", 
  "Technical Sergeant", "Technical Sergeant", "Petty Officer First Class",
  "E7", "Sergeant First Class", "Chief Petty Officer",
  "Gunnery Sergeant", "Master Sergeant OR First Sergeant", "Master Sergeant", 
  "Chief Petty Officer",
  "E8", "First Sergeant OR Master Sergeant", "Senior Chief Petty Officer", 
  "First Sergeant OR Master Sergeant", "Senior Master Sergeant OR First Sergeant", 
  "Senior Master Sergeant", "Senior Chief Petty Officer",
  "E9", "Sergeant Major OR Command Sergeant Major", 
  "Master Chief Petty Officer OR Fleet/Command Master Chief Petty Officer", 
  "Sergeant Major OR Master Gunnery Sergeant", 
  "Chief Master Sergeant OR First Sergeant", 
  "Chief Master Sergeant", 
  "Master Chief Petty Officer OR Fleet/Command Master Chief Petty Officer"
)

#join and clean up naming
armed_forces_ranked <- armed_forces %>%
  left_join(rank_map, by = "Pay_Grade") %>%
  mutate(
    Rank = case_when(
      Branch == "Army" ~ Army,
      Branch == "Navy" ~ Navy,
      Branch == "MarineCorps" ~ MarineCorps,
      Branch == "AirForce" ~ AirForce,
      Branch == "SpaceForce" ~ SpaceForce,
      Branch == "CoastGuard" ~ CoastGuard,
      TRUE ~ NA_character_
    )
  ) %>%
  select(Gender, Branch, Rank)
```

## Question 4: Visualization for the Armed Forces

In this section I created a two-way frequency table to see the impact of sex and rank in the US Armed Forces.

```{r}
freq_table <- armed_forces_ranked %>%
filter(Branch == "Army", !is.na(Rank)) %>% #filter out the N/A
group_by(Gender, Rank) %>% # group the gender and rank together
summarise(Count = n(), .groups = "drop") %>%
group_by(Gender) %>%
mutate(RelFreq = Count / sum(Count)) #calculate relative frequency

kable(freq_table, caption = "Army Enlisted Frequency Table: Gender by Rank")
```

## Question 5: Narrative Text for the Armed Forces Section

The visualization shows how gender and rank are distributed among enlisted people. Most men and women are revolving around the lower enlisted ranks, especially the Private First Class and Corporal/Specialist. Women make up a smaller share of higher ranks like Sergeant First Class and Sergeant Major. This pattern shows that gender and rank are not independent, men tend to lean towards senior ranks, while women tend to lean to the lower and middle class ranks.

# Popularity of Baby

## Question 6: Code for the Popular Baby Names Project

```{r}
#load libraries
library(babynames)
library(dplyr)
library(ggplot2)

#load baby names data set
data("babynames")

#find 10 most popular baby names for all years and sexes
top10_names <- babynames %>%
group_by(name) %>%
summarise(total = sum(n)) %>%
arrange(desc(total)) %>%
slice_head(n = 10)

#filter the data set to include only those names
top10_data <- babynames %>%
filter(name %in% top10_names$name)

#time series plot of name popularity
ggplot(top10_data, aes(x = year, y = n, color = name)) +
geom_line(linewidth = 1.1, alpha = 0.8) +
facet_wrap(~sex) +
labs(
title = "Top 10 Baby Names in the U.S. Over Time",
x = "Year",
y = "Number of Babies Given Name",
color = "Name"
) +
theme_minimal(base_size = 13) +
theme(
plot.title = element_text(face = "bold", hjust = 0.5),
plot.subtitle = element_text(hjust = 0.5, color = "gray30"),
legend.position = "right",
panel.grid.minor = element_blank()
)
```

## Question 7: Visualization for Popular Baby Names Project

This visualization shows the most common baby names in the United States over time. It compares how certain names rise and fall in popularity for boys and girls, showing cultural and social influences over different decades.

```{r fig.cap="Trends in the Top 10 U.S. Baby Names Over Time", fig.alt="Two line charts showing the number of babies with each of the top 10 names from 1880 to 2020, separated by sex."}
#time series plot of name popularity
ggplot(top10_data, aes(x = year, y = n, color = name, linetype = name)) +
  geom_line(linewidth = 1.1, alpha = 0.8) +
  facet_wrap(~sex) +
  labs(
    title = "Top 10 Baby Names in the U.S. Over Time",
    x = "Year",
    y = "Number of Babies Given Name",
    color = "Name",
    linetype = "Name"
  ) +
  scale_color_brewer(palette = "Dark2") + 
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray30"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Question 8: Narrative Text for the Popular Baby Names Project
The time plot shows the changing popularity of the 10 most common names across 120 years. Some names were common, then dropped off, while other stayed high for most of it. These different shifts show how culture and society can affect people's baby naming for new parents.

# The Box Problem

## Question 9: Code for the Box Problem

```{r}
#volume box
volume_box <- function(x) {
  V <- x * (48 - 2*x) * (36 - 2*x)
  return(V)
}

#load libraries
library(ggplot2)

#create the ggplot
box_plot <- ggplot(data.frame(x = c(0, 18)), aes(x = x)) +
  stat_function(fun = volume_box, color = "steelblue", linewidth = 1.2) +
  labs(
    title = "Volume of the Box vs. Cut Size",
    x = "Cut size x (inches)",
    y = "Volume (cubic inches)"
  ) +
  theme_minimal(base_size = 14)

#add alt text
box_plot + labs(
  alt = "Line graph showing how the volume of a box changes as cut size 
  increases. The curve rises, peaks near x = 6 in, then falls back to zero 
  as cuts become too large."
)
```

## Question 10: Narrative Text for the Box Problem

The plot shows a volume increase until the cut size reaches around 6 inches with a maximum volume around 5200 cubic inches. After the maximum the volume decreases because the dimensions shrink too much.

# Self-Reflection

Throughout this course, I have learned how to use data visualizations and stat tools in RStudio to explore real-world data. I have become very familiar with R and a lot of features within it, even though I am taking other 400 level R classes. Having good visuals that follow a good design from Tufte and Kosslyn is a necessity. I have also learned the importance of commenting sections of code in order to show what each block does. Overall, I developed good skills so far and want to keep advancing that.
